---
title: "Project1 Statistical Computation"
author: "Eyal Grinberg & Yam Rozen"
date: "2023-04-01"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

Q1 - Ladders & Snakes

An auxiliary function for summing digits.
```{r}
# the function receives a number and returns the final sum of it's digits. 

calc_digits_sum <- function(num) {
  digits <- as.numeric(strsplit(as.character(num), "")[[1]]) # from chatGPT: the function splits the number into its individual digits using the strsplit() function. It then converts each digit to a numeric type using the as.numeric() function.
  curr_sum <- sum(digits)
  if (curr_sum > 9) { # there's need to sum digits again
    return(calc_digits_sum(curr_sum)) #recursion
  }
  return(curr_sum)
}
```

data structures:
```{r}
# for checking ladders
base_of_a_ladder_vec <- c(2,7,8,15,21,28,36,51,71,78,87)
top_of_a_ladder_vec <- c(38,14,31,26,42,84,44,67,91,98,94)
# for checking snakes
head_of_a_snake_vec <- c(16,46,49,62,64,74,89,92,95,99)
tail_of_a_snake_vec <- c(6,25,11,19,60,53,68,88,75,80)
```

An auxiliary function for checking if a square is a head of a snake or a base of a ladder.
```{r}
# the function receives a vector and a number and returns the index of the number in the vector if it exists, -1 otherwise.

find_index_in_vec <- function(vec, num) {
  index <- which(vec == num)
  if(length(index) == 0) {
    return(-1)
  }
  return(index)
}
```

A function for one turn in the game.
```{r}
# receives the starting square as an argument.
# returns the end square.

perform_one_turn_Q1 <- function(start_square) {
  dice_roll <- sample(1:6, 1)
  mid_square <- dice_roll + start_square 
  # check for exceeding 100
  if (mid_square > 100) {
    mid_square <- 100 - (mid_square - 100)
  }
  # check sum digits condition
  if (calc_digits_sum(mid_square) == dice_roll) {
    mid_square <- floor(mid_square / 2)
  }
  # check for a snake
  index <- find_index_in_vec(head_of_a_snake_vec, mid_square)
  if (index != -1) {
    mid_square <- tail_of_a_snake_vec[index]
  }
  # check for a ladder
  index <- find_index_in_vec(base_of_a_ladder_vec, mid_square)
  if (index != -1) {
    mid_square <- top_of_a_ladder_vec[index]
  }
  # at this point 'mid_square' is actually the end square of this turn.
  return(mid_square)
}
```

the main function for performing a full game and counting the number of dice rolls.
```{r}
perform_one_game_Q1 <- function() {
  cnt <- 0
  start_square <- 1
  end_square <- -1
  while (end_square != 100) {
    end_square <- perform_one_turn_Q1(start_square)
    start_square <- end_square
    cnt <- cnt + 1
  }
  return (cnt)
}
```

auxiliary function for CI calculation:
```{r}
calc_CI <- function(data, conf) {
  alpha <- 1 - conf
  q_z <- qnorm(1 - alpha/2)
  n <- length(data)
  avg <- mean(data)
  sd <- sd(data)
  upper <- avg + q_z * (sd / sqrt(n)) 
  lower <- avg - q_z * (sd / sqrt(n)) 
  CI <- c(lower, upper)
  names(CI) <- c("Lower", "Upper")
  return (CI)
}
```

5000 games simulation:
```{r}
sim_results_Q1 <- replicate(5000, perform_one_game_Q1())
```

Mean and CI calculations:
```{r}
round(mean(sim_results_Q1), 3) # 3 digits after the decimal point as requested
```

```{r}
round(calc_CI(sim_results_Q1, 0.95), 3)
```



Q2 - Tic Tac Toe

```{r}
# auxiliary function that checks if any of the players won after a move was made.
# the input is the the last square that was played by one of the players as well as the board and it's size and the player.
# the function returns 1 if X player has won.
# the function returns 2 if O player has won.

check_win <- function(player, board, size, square) {
  # extract the relevant row and col number
  row_index_of_square <- ceiling(square / size)
  col_index_of_square <- -1
  modulu_res <- square %% size
  if (modulu_res) {
    col_index_of_square <- modulu_res
  }
  else { # the column of square is a non-remainder multiple of size ---> should be the last column in the matrix.
    col_index_of_square <- size
  }
  
  col_match <- row_match <- primary_diag_match <- secondary_diag_match <- 0
  # check rows and columns
  for (i in 1:size) {
    if (row_match == -1 && col_match == -1) {break} # there's no chance for a row/col match, stop searching.
    if (row_match != -1) { # a little optimization, if we found a square in the row that doesn't match, we won't keep checking that row. 
      if (board[row_index_of_square, i] == player) {row_match <- row_match + 1}
      else {row_match <- -1}
    }
    if (col_match != -1) { # same optimization for columns
      if (board[col_index_of_square, i] == player) {col_match <- col_match + 1}
      else {col_match <- -1}
    }
  }
  
  # if the square is not on any of the diagonals, we only need to check the row and the col of the given square. otherwise we need to check also the diagonals.
  if (row_index_of_square == col_index_of_square || 
      col_index_of_square == size - row_index_of_square + 1) { # check if the square is on one of the two diagonals.
  # same optimization for diagonals too.
    for (i in 1:size) {
      if (primary_diag_match == -1 && secondary_diag_match == -1) {break}
      if (primary_diag_match != -1) { 
        if (board[i, i] == player) {primary_diag_match <- primary_diag_match + 1}
        else {primary_diag_match <- -1}
      }
      if (secondary_diag_match != 1) {
        if (board[i, size - i + 1] == player) {secondary_diag_match <- secondary_diag_match + 1}
        else {secondary_diag_match <- -1}
      }
    }
  }
  
  if (row_match == size || col_match == size ||
      primary_diag_match == size || secondary_diag_match == size) {
    return (player) # should return 1 or 2, depends on which player played this turn
  }
  return(-1)
}
```

One game function
```{r}
# the function returns 1 if X player has won, 2 if O player has won, and 0 if it's a tie

perform_one_game_Q2 <- function(size) {
  
  board <- matrix(-1, size, size)
  permuted_squares <- sample(c(1:(size^2))) # here all the "magic" happens, we created a permutation of the squares and each turn take a square sequentially from that vector instead of sampling in each iteration.
  
  i <- 1
  while (i <= size^2) {
    # X player's turn
    board[ permuted_squares[i] ] <- 1
    if (i >= 2 * size) { # need to check win condition only after size turns were played by each player 
      win_check_res <- check_win(1, board, size, permuted_squares[i])
      if (win_check_res != -1) { # game over
        return(win_check_res)
      }
    }
    i <- i + 1
    if (i == size^2 + 1) {break} # if there's an odd number of squares the X player will play one more turn.
    # O player's turn
    board[ permuted_squares[i] ] <- 2
    if (i >= 2 * size) { # need to check win condition only after 'size' turns were played 
      win_check_res <- check_win(2, board, size, permuted_squares[i])
      if (win_check_res != -1) { # game over
        return(win_check_res)
      }
    }
    i <- i + 1
  }
  # otherwise it's a tie
  return(0)
}
```

simulations:
```{r}
sim_size <- 5000
# matrix initialization
prob_mat <- matrix(nrow = 3, ncol = 15)
row.names(prob_mat) <- c("Win probability X","Win probability O","Tie probability")
colnames(prob_mat) <- c('3x3','4x4','5x5','6x6','7x7','8x8','9x9','10x10','11x11','12x12','13x13','14x14','15x15','25x25','50x50')
# performing simulations
for (i in 3:15) {
  results_game_size_i <- replicate(sim_size, perform_one_game_Q2(i))
  prob_mat[1,i-2] = mean(results_game_size_i == 1) # games that X has won returned 1
  prob_mat[2,i-2] = mean(results_game_size_i == 2) # games that X has won returned 2
  prob_mat[3,i-2] = mean(results_game_size_i == 0) # games that X has won returned 0
}
for (i in 1:2) {
  results_game_size_i <- replicate(sim_size, perform_one_game_Q2(i*25))
  prob_mat[1,i+13] = mean(results_game_size_i == 1) # games that X has won returned 1
  prob_mat[2,i+13] = mean(results_game_size_i == 2) # games that X has won returned 2
  prob_mat[3,i+13] = mean(results_game_size_i == 0) # games that X has won returned 0
}
prob_mat
```

```{r}
board_size <- c(3:15,25,50)
plot(x = board_size, y = prob_mat[1,], xlab = "Board size",
     ylab = "Estimated win probability of X", col = "red", main = "Player X",lwd = 3, type = "b")
plot(x = board_size, y = prob_mat[2,], xlab = "Board size",
     ylab = "Estimated win probability of O", col = "blue", main = "Player O",lwd = 3, type = "b")
plot(x = board_size, y = prob_mat[3,], xlab = "Board size",
     ylab = "Estimated Tie probability", col = "orange", main = "Ties plot",lwd = 3, type = "b")
```

As shown in the plots, it seems that as the board size increases the tie probability increases too, and the probabilities that any of the two players will win go down towards 0.
the intuition for it is that on a larger board the majority of the square permutations will end in a tie.


Bonus Section - performing 5,000,000 simulations
uncomment to run the simulations
```{r}
# # Set the number of simulations
# n_sims <- 5000000
# sim_result_size_25 <- vector("numeric", n_sims)
# sim_result_size_50 <- vector("numeric", n_sims)
# 
# # Iterate over the simulations
# for (i in 1:n_sims) {
#   # Run the function for each simulation
#   sim_result_size_25[i] <- perform_one_game_Q2(25)
#   sim_result_size_50[i] <- perform_one_game_Q2(50)
# 
#   # Check if the current iteration is a multiple of 50,000
#   if (i %% 50000 == 0) {
#     # Print a message to indicate the progress of the simulation
#     cat(i / 50000,"% of the simulations completed. \n")
#   }
# }
# ```
# Probabilities
# ```{r}
# # matrix initialization
# prob_mat <- matrix(nrow = 3, ncol = 2)
# row.names(prob_mat) <- c("Win probability X","Win probability O","Tie probability")
# colnames(prob_mat) <- c('25x25','50x50')
# 
# # performing simulations
# prob_mat[1,1] = mean(sim_result_size_25 == 1) # games that X has won returned 1
# prob_mat[2,1] = mean(sim_result_size_25 == 2) # games that X has won returned 2
# prob_mat[3,1] = mean(sim_result_size_25 == 0) # games that X has won returned 0
#   
# prob_mat[1,2] = mean(sim_result_size_50 == 1) # games that X has won returned 1
# prob_mat[2,2] = mean(sim_result_size_50 == 2) # games that X has won returned 2
# prob_mat[3,2] = mean(sim_result_size_50 == 0) # games that X has won returned 0
# 
# prob_mat
```

Plots
```{r}
# board_size <- c(25,50)
# plot(x = board_size, y = prob_mat[1,], xlab = "Board size",
#      ylab = "Estimated win probability of X", col = "red", main = "Player X",lwd = 3, type = "b")
# plot(x = board_size, y = prob_mat[2,], xlab = "Board size",
#      ylab = "Estimated win probability of O", col = "blue", main = "Player O",lwd = 3, type = "b")
# plot(x = board_size, y = prob_mat[3,], xlab = "Board size",
#      ylab = "Estimated Tie probability", col = "orange", main = "Ties plot",lwd = 3, type = "b")
```
As expected, for 5,000,000 simulations the estimators for the win probabilities of any player is 0 and for the tie probability is 1. 
